---
title: "Adapter reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adapter reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(cellspecR)
```

`cs_formats()` is the machine-readable index of supported reader grammars.
Each adapter preserves the source column name in `dictionary$source_name` and
records its adapter and version in provenance.

```{r}
cs_formats()
```

The QuPath adapter accepts both channel-first headers such as `DAPI: Mean` and
compartment-first headers such as `Nucleus: CD3e: Mean`. The tiled adapter
adds an ownership pass over outer tile boxes; halo rows are counted and
dropped. MCQuant and segmantR readers treat `centroid_col` as x and
`centroid_row` as y, and require pixel calibration.

When no grammar is decisive, `cs_read(format = "auto")` returns a classified
ambiguity error and lists the candidates. This prevents a plausible but wrong
column match from reaching downstream analysis.
