---
title: "Reading cell measurement exports"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading cell measurement exports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(cellspecR)
```

Readers are selected explicitly when a project contains more than one
possible grammar. The bundled synthetic export uses QuPath-style headers.

```{r}
path <- system.file("extdata", "qupath-0.7.0-instanseg.tsv", package = "cellspecR")
cs_detect_format(path)
x <- cs_read(path, format = "qupath")
print(x)
cs_validate(x)
```

For an unrecognised table, map the source columns and describe measurements in
the dictionary at the boundary.

```{r}
map <- cs_column_map(
  cell_id = "Cell ID", x = "Centroid X um", y = "Centroid Y um",
  measurements = data.frame(
    source_name = "CD3e: Mean", kind = "intensity", marker = "CD3e",
    compartment = "cell", statistic = "mean", unit = "a.u."
  )
)
map
```

Pixel coordinates always need a calibration. A missing or inconsistent unit
is an error rather than a guess.
