---
title: "The cellspec 1.0 format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The cellspec 1.0 format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(cellspecR)
cols <- cs_vocabulary("columns")
voc <- cs_vocabulary("vocabularies")
checks <- cs_vocabulary("checks")
column_table <- function(component) {
  tab <- cols[cols$component == component, c("column", "type", "requirement", "description")]
  tab$column <- paste0("`", tab$column, "`")
  rownames(tab) <- NULL
  knitr::kable(tab, col.names = c("Column", "Type", "Requirement", "Meaning"))
}
values <- function(vocabulary) {
  paste0("`", voc$value[voc$vocabulary == vocabulary], "`", collapse = ", ")
}
```

This document is the normative description of `cellspec`
`r cs_spec_version()`, the table format for segmented cells from multiplexed
tissue images. The key words **MUST**, **SHOULD** and **MAY** are used as in
RFC 2119. Every table below is generated from the objects that the validator
uses (`cs_vocabulary()`), so the text and the checks cannot drift apart.

This specification is shipped with cellspecR version
`r as.character(utils::packageVersion("cellspecR"))`.

## 1. Object

A `cellspec` object is a list of class `"cellspec"` with these components:

| Component | Type | Required | Content |
|---|---|---|---|
| `spec_version` | string | yes | `"`r cs_spec_version()`"` |
| `cells` | data frame | yes | one row per cell: identity and geometry (section 2) |
| `measurements` | double matrix | yes (may have 0 columns) | cells x features, rows in the order of `cells` (section 3) |
| `dictionary` | data frame | yes | one row per measurement column (section 4) |
| `images` | data frame | yes | one row per image (section 5) |
| `channels` | data frame | yes | one row per image and channel (section 5) |
| `provenance` | list | yes | origin of the data (section 6) |
| `adjacency` | data frame or `NULL` | no | cell-cell contacts (section 7) |

A whole slide carries up to about 1,000 numeric features for millions of
cells. A double matrix stores them compactly and maps directly to an assay of
a `SpatialExperiment`; `cells` stays a data frame because it holds mixed
types. Accessors return plain data frames. Rows of `measurements` are aligned
with `cells` by position; row names are not used.

```{r}
x <- cs_example()
x
```

## 2. Cells

```{r, echo = FALSE}
column_table("cells")
```

Rules:

- The key `(image_id, cell_id)` MUST be unique.
- `x` and `y` MUST be finite. `area`, when present, MUST be positive or `NA`.
- Identifiers are character columns, never row names.
- Additional columns are allowed and SHOULD be snake_case. The prefixes
  `region__` and `boundary_distance__` are reserved for region annotations
  and `cs__` for internal use. A layer name MUST match
  `^[A-Za-z0-9][A-Za-z0-9_.-]*$`.
- `cells` MUST NOT contain clinical variables; `subject_id` belongs in
  `images`.

## 3. Measurements

- A double matrix with one row per cell.
- Column names are feature identifiers and MUST equal
  `dictionary$feature_id`, in the same order.
- Values are kept as exported by the source tool, in raw units. No
  transformation is applied on import. `NaN` and infinite values from sources
  are converted to `NA` and counted in `provenance$counts`.

## 4. Dictionary

```{r, echo = FALSE}
column_table("dictionary")
```

Controlled vocabularies:

- `kind`: `r values("kind")`.
- `compartment`: `r values("compartment")`, or `NA` for `other` features.
- `statistic` for intensity features: `r values("intensity_statistic")`.
- `statistic` for shape features: `r values("shape_statistic")`.
  Unknown shape measurements are stored as `kind = "other"`, never dropped.
- `unit`: `r values("unit")`.

Feature identifiers are generated by readers and follow one rule per kind:

| Kind | Identifier | Example |
|---|---|---|
| intensity | `<compartment>:<marker>:<statistic>` | `nucleus:FOXP3:mean` |
| shape | `<compartment>:<statistic>` | `cell:area` |
| other | `other:<name>` | `other:detection_probability` |

Downstream code MUST select features through their dictionary columns, for
example with `cs_features()`, and not by splitting identifiers. Marker names
MUST be trimmed and MUST NOT contain `:`. Because identifiers follow the rule
above and are unique, no two features can describe the same kind, marker,
compartment and statistic.

```{r}
cs_features(x, kind = "intensity", marker = "FOXP3")
```

## 5. Images and channels

```{r, echo = FALSE}
column_table("images")
```

```{r, echo = FALSE}
column_table("channels")
```

Every intensity marker in the dictionary MUST appear in `channels$marker` for
every image of the object. A channel without measurements is allowed, for
example a nuclear stain used only for segmentation.

## 6. Provenance

`provenance` is a list with at least these fields:

| Field | Content |
|---|---|
| `producer` | `tool` and `version` of the software that measured the cells |
| `reader` | `package`, `version`, `adapter` and `adapter_version` that built the object |
| `created_utc` | creation time, `YYYY-MM-DDTHH:MM:SSZ` |
| `inputs` | one entry per input file: `path_basename`, `bytes`, `sha256` |
| `counts` | `rows_read`, `rows_kept`, `na_converted`, `nonfinite_converted` |
| `parameters` | reader parameters, e.g. where the pixel size came from |
| `history` | steps that changed content: `step`, `function`, `time_utc`, `details` |

Input paths are stored as file names plus hash. Full paths leak private
directory structures and are stored only when a user explicitly asks for it.

## 7. Adjacency (optional)

```{r, echo = FALSE}
column_table("adjacency")
```

`method` is one of `r values("adjacency_method")`. Each unordered pair of
cells appears at most once, ordered so that `cell_id_a` sorts before
`cell_id_b` in byte order, and both cells MUST exist in `cells`. Subsetting an
object drops contacts whose cells were removed.

## 8. Coordinate frame and units

- The origin is the top-left corner of the top-left full-resolution pixel;
  x increases to the right and y increases downward (the QuPath convention).
- The centre of the pixel with 0-based index `i` lies at
  `(i + 0.5) * pixel_size`.
- Centroids `x` and `y` are in micrometres. Readers that receive pixel
  centroids MUST know the pixel size, from the file or from an argument, and
  MUST stop otherwise.
- When a tool reports both micrometre and pixel centroids, readers check that
  `abs(x - x_px * pixel_size) <= 0.5 * pixel_size` and stop on disagreement.
- The micro sign appears in exports as U+00B5, U+03BC, `um` or the mis-decoded
  `µ`. Readers normalise these variants before matching headers and keep the
  original header in `source_name`.

## 9. Directory layout

A `cellspec` directory holds one object on disk:

```
<dir>/
  cells.parquet       cells columns, then one column per feature_id
  cellspec.json       spec_version, column types, dictionary, images,
                      channels, provenance
  adjacency.parquet   optional
  MANIFEST.sha256     "<sha256>  <relative path>" for every file above
  DONE                written last; holds the SHA-256 of MANIFEST.sha256
```

- The cell table is stored as Parquet by default, which keeps every double
  bit-exact. A tab-separated alternative (`cells.tsv.gz`,
  `adjacency.tsv.gz`) is readable by any tool; its numbers carry 15
  significant digits, so values can differ from the originals in the last
  representable digits.
- `cellspec.json` validates against the JSON Schema shipped in
  `system.file("schema", "cellspec-1.0.0.schema.json", package = "cellspecR")`.
- Writers stage the directory under a temporary name and rename it at the
  end, so a directory without `DONE` was not completely written.
- Readers refuse a directory without `DONE`, with a manifest mismatch, or with
  an unsupported major `spec_version`.

## 10. Validation rules

`cs_validate()` checks an object against the rules above. Structure checks
implement the MUST rules and fail; semantic checks flag implausible content
and warn.

```{r, echo = FALSE}
tab <- checks
tab$check <- paste0("`", tab$check, "`")
knitr::kable(tab, col.names = c("Check", "Level", "Rule"))
```

## 11. Versioning

- `spec_version` follows semantic versioning. Minor versions only add
  optional columns or components.
- A reader of major version 1 accepts every 1.y object whose minor version is
  not newer than its own.
- Readers of tool exports carry their own `adapter_version`, which changes
  whenever their output for the same input changes.
